### 2023-09-01 sections/1-introduction.tex
changes for clarity

\section{Introduction}
\label{sec:introduction}
Often dubbed "the Big Data explosion," rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
This is especially true in the field of astronomy. 
The Sloan Digital Sky Survey (SDSS)~\cite{blanton2017sdss}, for example, in its APOGEE~\cite{alam2015eleventh} dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. 
Their MaNGA dataset also contains spectral measurements across the face of each of about 10,000 galaxies across 2,700 deg$^2$. [add sentence about big ANN competition]

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters.

More recently, Big-ANN launched NeurIPS'23: Billion Scale Approximate Nearest Neighbor Search Challenge, 
which aims to encourage the development of indexing data structures and search algorithms for practical variants of Approximate Nearest Neighbor (ANN) or Vector search.
This challenge tests submissions' performances with the YFCC 100M~\cite{johnson2019billion} dataset, the Yandex Text-to-Image 10M~\cite{diskann-github}, cross-modal dataset, 
the MSMARCO passage retrieval dataset~\cite{Bruch2023AnAA}, and a 10M slice of the MS Turing data set~\cite{Singh2021FreshDiskANNAF}. These datasets 
are 192, 200, 30,000, and 100 dimensional respectively.

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating 
that computer architecture improvements will not "catch up" to our computational needs in the near future. As a result, analysis of these large datasets 
requires better algorithms.

@@

In particular, $k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}; 
naive implementations of kNN, whose time complexity is linear in the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not 
exploit the geometric and topological structure inherent in big datasets. Approximate search is sufficient for some applications, but the 
need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$.

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{indyk1999sublinear}, 
clever indexing techniques such as the FM Index~\cite{simpson2010efficient}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the geometric and topological structure inherent in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CAKES (CLAM-Accelerated $K$-nearest-neighbor 
Entropy-scaling Search). CLAM (Clustered Learning of Approximate Manifolds), developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the CHESS clustering algorithm. 
Using the cluster tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform 
$k$-nearest neighbors search in addition to $\rho$-nearest neighbors search (also known as ranged nearest neighbors search).

This paper focuses on four novel accelerated $k$-nearest neighbors search algorithms. We explain some 
implementation improvements to CHESS, as well as improved genericity across distance functions. 
We also include proof of some geometric guarantees for clusters. Finally, we provide a comparison of 
CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion} on several datasets 
from the ann and big ann benchmarks suites.

>>>>

\section{Introduction}
\label{sec:introduction}
Often dubbed "the Big Data explosion," rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
This is especially true in the field of astronomy. 
The Sloan Digital Sky Survey (SDSS)~\cite{blanton2017sdss}, for example, in its APOGEE~\cite{alam2015eleventh} dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. 
Their MaNGA dataset also contains spectral measurements across the face of each of about 10,000 galaxies across 2,700 deg$^2$.

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters; 
the GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences.

More recently, Big-ANN's NeurIPS'23: Billion Scale Approximate Nearest Neighbor Search Challenge has 
solicits development of data structures and search algorithms for Approximate Nearest Neighbor (ANN) search
on four massive datasets: YFCC 100M~\cite{johnson2019billion} dataset, the Yandex Text-to-Image 10M~\cite{diskann-github}, cross-modal dataset, 
the MSMARCO passage retrieval dataset~\cite{Bruch2023AnAA}, and a 10M slice of the MS Turing data set~\cite{Singh2021FreshDiskANNAF}. These datasets 
are 192, 200, 30,000, and 100 dimensional respectively, and submissions are evaluated with 
a ten million point sample of each.

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating 
that computer architecture will not "catch up" to our computational needs in the near future. As a result, analysis of these large datasets 
requires better algorithms.

@@

In particular, $k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Naive implementations of kNN, whose time complexity is linear in the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not 
exploit the geometric and topological structure inherent in large datasets. Approximate search is sufficient for some applications, but the 
need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$.

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{indyk1999sublinear}, 
clever indexing techniques such as the FM Index~\cite{simpson2010efficient}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the geometric and topological structure inherent in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CAKES (CLAM-Accelerated $K$-nearest-neighbor 
Entropy-scaling Search). CLAM (Clustered Learning of Approximate Manifolds), developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the CHESS clustering algorithm. 
Using the cluster tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform
$k$-nearest neighbors search.

This paper focuses on four novel exact $k$-nearest neighbors search algorithms. We also explain some 
implementation improvements to the clustering and $\rho$ nearest neighbors search algorithms in CHESS, 
as well as improved genericity across distance functions. Additionally, we prove a guarantee about the scaling behavior of cluster radii. 
Finally, we provide a comparison of 
CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion} on several datasets 
from the ANN and Big ANN benchmarks suites.

### 2023-09-01 sections/2-methods.tex
changes for clarity

We define a \emph{distance function} $f : (\textbf{X}, \textbf{X}) \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
When, in addition to these constraints, the distance function obeys
the triangle inequality, it is a \emph{distance metric}, and in such cases we can guarantee that our search algorithms have perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point in the dataset, 
which we define as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}~\cite{fefferman2016testing}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=2.5in]{images/lfd_plots/lfd_plot_apogee.png}
    \caption{Mean fractal dimension of APOGEE clusters as a function of
    depth when clustered based on L2 norm. Each plot line represents a distinct
    decile of fractal dimension. Beyond a depth of 56, no clusters are further
    divided due to the minimum-cluster-cardinality stopping criteria.}
    \label{fig:methods:lfd_apogee}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics[width=2.5in]{images/lfd_plots/lfd_plot_greengenes.png}
    \caption{Mean fractal dimension of GreenGenes clusters as a function of
    depth when clustered based on Hamming distance. Each plot line represents
    a distinct decile of fractal dimension.}
    \label{fig:methods:lfd_greengenes}
\end{figure}

Figure~\ref{fig:methods:lfd_apogee} shows the mean local fractal dimension of the APOGEE dataset at each level of hierarchical clustering, 
under the L2 (Euclidean) distance metric. Each plot line represents a decile of fractal dimension; note that, other than the most extreme 10\% of clusters,
virtually all clusters have a local fractal dimension of less than $2 \ll 8,575$. This suggests that APOGEE is a good candidate for use with CAKES.

Figure~\ref{fig:methods:lfd_greengenes} shows the mean local fractal dimension of the GreenGenes dataset at each level of hierarchical clustering,
under the Hamming distance metric. Each plot line represents a decile of fractal dimension; once again, other than the most extreme 10\% of clusters,
virtually all clusters have a local fractal dimension of less than $2 \ll 2,250$. This suggests that GreenGenes is a good candidate for use with CAKES.

@@

With these concepts defined, we can now pose the $k$- and $\rho$- nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in X \setminus S_q$, $f(q, x) \leq \max\{f(y, q): y \in S_q \}$; that is,
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$; that is, 
find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

>>>>

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
When, in addition to these constraints, the distance function obeys
the triangle inequality, it is a \emph{distance metric}, and in such cases we can guarantee that our search algorithms have perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point 
and length scale in the dataset, 
which we define as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space is known as the \emph{manifold hypothesis}~\cite{fefferman2016testing}.

Figures 2 and 3 in~\cite{ishaq2019clustered} illustrate the low fractal dimensions of two datasets used with $\rho$-NN search in CHESS. 
We observe that for both APOGEE (Figure 2 in~\cite{ishaq2019clustered}) and GreenGenes (Figure 3 in~\cite{ishaq2019clustered}), other than the most extreme 10\% of clusters, virtually all clusters have a local fractal dimension of less than $2 \ll 8,575$
and $2 \ll 2,250$ respectively.
This suggests that APOGEE and GreenGenes are good datasets for use with entropy-scaling search algorithms like CHESS and CAKES.

@@

With these concepts defined, we can now pose the $k$- and $\rho$- nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \setminus S_q$, $f(q, x) \leq \max\{f(y, q): y \in S_q \}$; that is,
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$; that is, 
find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.
